Genetic Dilution erodes productivity: Exploring farmers' low adoption levels of improved maize in Ethiopia

### 2023-03-22 paper.tex
Updates from Overleaf

    \clearpage\maketitle
	\thispagestyle{empty}
	\vspace*{-2em}
    \begin{center}\begin{abstract}
			\noindent  
			Despite large investments in  developing new  maize varieties, evidence shows that few farmers consistently adopt this new technology over time. We study the adoption trajectories of farmers over a three-round panel data set to explain low adoption rates with heterogeneous returns to adoption. By relying on self-reported use of improved seeds, we find a puzzling answer for why farmers do not use new seed varieties: adoption leads to seemingly small and negative returns. Heterogeneity analysis across different levels of comparative advantage show that farmers who adopt do not have a comparative advantage in adopting improved seed, nor do they exhibit high returns to adopting. To reconcile this result with the expected higher yields for improved seeds, we exploit unique information on maize DNA-fingerprinting collected over the same areas in 2018. We find that the negative returns to adoption mask heterogeneity in the actual quality of the genetic material of the seeds farmers categorized as improved varieties in the self-reported data. Positive returns to adoption are found for those farmers that use higher-purity germplasm, drought-tolerant maize, and newly released varieties. When we account for misclassification in our heterogeneity analysis, we find that the puzzle is resolved. Our findings point to the wide dispersion of older and genetically diluted varieties, for which poorer farmers may be paying a premium. The implications of our findings speak to the need for policies to better target context and geography, expand accessibility of improved seeds, and make higher yields varieties more inclusive.

@@

        {\small \noindent\emph{JEL Classification}:
        	\emph{Keywords}: Improved maize, DNA-fingerprinting, Group Random Coefficient, Endogenous Switching Regression, Heterogeneous returns}

@@

%%% Why it is important
Food security is a major challenge in many East African countries, and Ethiopia is no exception, having historically struggled to provide an adequate and reliable food supply \citep{Ramakrishna2002-hv, Jaleta2018-oj}. Since the drought of 1984, Ethiopia has become increasingly reliant on maize, and as of today, maize constitutes the most widespread crop in the country.\footnote{FAO statistics show the total area cultivated with Maize in 2019 is 25\% larger than the next most widespread crop (sorghum), and 27\% and 139\% larger than the next two following crops, wheat and barley, respectively.} There was an upward trend in the total area cultivated with maize since 1990, with average yields almost doubling during this period. \footnote{Improved security conditions and maize productivity contributed with an important decrease in undernourishment which fell from 47\% in 1990 to 14\% in 2018 (The World Bank, 2021). Despite these positive gains in total area planted and average yields, food insecurity still affects more than half the population of Ethiopia.The country has also recently seen increases in food insecurity because of COVID-19, the conflict in Ethiopia’s Tigray region and the desert locust outbreak.} Despite progress a large productivity gap persists between the African continent and the rest of the developing world. From 2016 to the present, cereal crop yields in Sub-Saharan Africa have consistently been around 70-80\% lower than Europe or North America \citep{worldbankyield}. The persistence of this gap is often attributed to different adoption rates of green revolution technologies, which include improved seeds. Evidence has shown that the adoption of improved seed leads to increases in agricultural yields \citep{Carter2014-fm}, food security \citep{Shiferaw2014-op}, and poverty reduction \citep{Minten2008-tj}.\footnote{Nevertheless, critics of green revolution technologies have also claimed that these technological packages have not been as profitable for many farmers in Africa as promised. Even the most successful of these countries, Ethiopia, is not on track to meet the goals of the Alliance for a Green Revolution for Africa (AGRA). This is due to high costs, and heavy dependence on subsidies \citep{Wise20}.}

@@

Despite evidence and numerous technological breakthroughs in maize germplasm with increased yields and drought tolerance, only 13\% of maize farmers in Ethiopia have adopted improved seeds consistently between 2011 and 2015. A large percentage has adopted new varieties inconsistently, and the largest share of farmers (62\%) have not used improved seeds at all during the same period. Several answers have been proposed to explain such low adoption of improved varieties. These include lack of liquidity and other economic constraints \citep{Carter2014-fm}, growing conditions such as the type of soil, input use, and farmer characteristics \citep{Munshi2004-og}, imperfections in credit markets \citep{Croppenstedt2003-pq}, ambiguously defined property rights \citep{Place2000-el}, lack of commitment devices \citep{Duflo2009-iv}, high transportation costs that increase the cost of agricultural inputs \citep{Byerlee2013-qk}, and improved seed supply constraints \citep{Bird2020-nt}. Additionally, the net returns to adoption of improved seeds are often lower than expected in the long term because a large percentage of farmers may  use diluted improved seed varieties, or ones that have been recycled over many generations. This is a concern as there is evidence that yield improvement and disease resistance wanes over time \citep{Abate2015-rj}. The variety of these constraints and their dispersion across subpopulations and locations speaks to considerable heterogeneous returns to adoption.

@@

In the absence of a natural experiment or randomized intervention, it is difficult to properly estimate the effect of adoption as there is potential selection of certain farmers into adoption improved varieties. This selection bias is more pervasive when the goal is to explore heterogeneous effects of adoption. Alternatively, one can make assumptions about the structure of the underlying error distribution to properly identify returns to adoption. In this paper, we revisit the latter mechanism proposed by \cite{Suri2011-oi}, where a lack of adoption of improved seed varieties is the result of farmers' heterogeneous returns to the new technology. Ignoring such heterogeneity hides the true impact of adoption, as farmers with different characteristics and skill levels observe different returns from adoption despite the technology providing high average returns. The heterogeneous returns to adoption are modeled as a farmer's unobservable comparative advantage. In the original paper, \cite{Suri2011-oi} finds that farmers’ lack of adoption is driven by insufficient net benefits to the technology stemming from poor infrastructure that increases the cost of access. Similar work in Ethiopia on improved chickpea varieties suggests that the economic returns to adoption plays a major role in the decision to adopt. \citep{Michler2018-wk}.

@@

Our results are divided into two parts. We start by using the first three survey rounds of the Ethiopian Socioeconomic Survey (ESS), which is a panel  covering  2011, 2013, and 2015. For each round, farmers self-reported on the seed variety they used (improved or traditional). By using the adoption trajectories over three rounds, we find the returns to adoption are in most cases, not statistically significant, and only significant for adoption in the third wave of the survey. This posits a puzzle that motivates our next estimation strategy.

Our second set of results use the fourth round of the ESS (conducted in 2018), which collected DNA fingerprints of harvested cropcuts in selected plots and compared its genetic material with a reference library of available varieties in Ethiopia, hence, providing a unique opportunity to understand why improved seed varieties seem not to increase returns when adopted. These data provides critical information not only about the seed types but also about their purity, sources, and their year of release. We exploit this uniquely rich information to compare returns of verifiable improved seed varieties against self-reported use of new, second generation or recycled varieties. For this, we leverage different purity thresholds (or degrees of no-contamination with other varieties), year of release, types of varieties, and their sources. Because the latest survey round is not a continuation of the first three rounds, our second estimation strategy implements an Endogenous Switching Regression (ESR) approach, a standard estimation strategy to study the question of adoption with cross-sectional data (\citealt{Marenya2020-kb}, \citealt{Shiferaw2014-op}), given its capacity to account for unobserved heterogeneity in adoption. The results of the ESR show negative effects of adopting improved seeds, as per farmer self-reports. But the comparison of different improved seed definitions using the DNA fingerprinting shows that when the definition of improved seed is based on greater purity of the germplasm (95\% of purity or more), the returns to adoption for those that used such improved varieties becomes positive and significant. We obtain similar results when evaluating the specific effects of adopting drought-tolerant maize varieties, CGIAR or Exotic varieties, or more recently released varieties.

Taken together, these results offer a plausible answer to the puzzle of low adoption levels of new varieties. The negative and not statistically significant returns to adoption we find from both estimation strategies reflect the loose definition of improved seeds when such improved varieties are self-reported by farmers. Only a subset of farmers use the purest genetic varieties, drought-tolerant varieties, or newly released varieties, which provide the highest returns. This implies that poorer farmers are paying a premium for genetically diluted hybrids, which has far-reaching policy implications about the need to foster a more inclusive strategy when making the best performing seeds available to all farmers.

Finally, to integrate our results together, we use the misclassification rates from the 4th round of the data and re-estimate the GRC results to be robust to misclassification. This method takes inspiration from \cite{michuda2021three} to account for the fact that the self-reports from the first three rounds of the ESS data may be misclassified and attenuated. We find that once we account for misclassification, heterogeneity in adoptions becomes very similar to other contexts: higher comparative advantage is associated with lower returns, pointing to significant barriers to adoption for those that would benefit the most from it. Incorporating misclassification rates from the fourth round of the ESS into the GRC estimation resolves the puzzle posed by the GRC results.

Our results contribute to the literature on the drivers of low adoption rates of improved seed varieties. The consensus in this literature is that the success of improved varieties depends on a range of factors, and that the widespread adoption of  improved seeds is the result of complementary conditions and inputs that allow farmers to extract the full potential of these varieties. Specifically, extension services and on-farm field trials, seed variety characteristics and rainfall have been found to be crucial to the adoption of improved maize in Tanzania \citep{Kaliba2000-jh}. In the specific case of Ethiopia, labor, fertilizer use, farmers’ experience with extension packages, rainfall suitability and prices have been shown to increase the opportunity cost of not adopting improved wheat varieties \citep{Wale2006-bv}. Market access and the accessibility of extension services are also critical in the adoption of improved chickpea varieties in Ethiopia \citep{Verkaart2019-ol}. Other important determinants of heterogeneous effects include risk aversion \citep{Holden2016-vy}, farm size \citep{Ghimire2015-bd}, and credit constraints \citep{Simtowe2008-jn,Balana2020-hx}.

@@

We contribute to this literature by exploring heterogeneous comparative advantages to adoption in Ethiopia between 2011 to 2015. %, controlling for important drivers of adoption such as rainfall (found to be the main driver of heterogeneous effects for adopting agronomic packages \citealt{Marenya2020-kb}, \citealt{Katengeza2019-af}), labor and input use. 
Our findings provide an alternative hypothesis to the puzzle of low maize adoption, namely that the wide dispersion of older generations and recycled improved seeds may be resulting in lower yields and altering farmers' input use. The null and negative effects that we see in the GRC results are actually stemming from misclassification in self-reported adoption status. This paper is the first to take seriously the bias that this can create when estimating the treatment effect of adoption, both in a homogeneous and heterogeneous setting. This also implies a methodological discovery when investigating hybrid seed adoption: the absence of a true control group to use for analysis. With more time and more genetic dilution, it becomes more likely that even those that do report hybrid use are still likely using some level of hybrid.

@@

There were significant differences between the analysis carried out in ESS 1-3 and the most recent, ESS4. Most notably, the ESS4 is no longer a panel dataset made up of the households of ESS 1-3, but it includes important information on the DNA fingerprinting of seed type, which can be used to calculate misclassification rates. This is important information as the first three waves of ESS only included self-reported hybrid maize status. Since our estimation strategy to calculate returns to adoption requires a balanced panel dataset of maize growers, our main analysis includes the first three waves of the ESS. We use the fourth wave to provide further light on the low levels of returns from adoption we find in the first three ESS rounds, and to explore the impact of specific types of improved seeds.

@@

Table \ref{tbl:summary} shows a set of summary statistics for the sample. The mean parcel size for a household is about 0.15 Ha, with a maximum size of less than 3.7 Ha. Harvest labor is disproportionately skewed to family participation with around 20 days of harvest labor being conducted by family members and only about 2 days for hired labor. This is indicative of either \textbf{malfunctioning} labor markets, income constraints or high supervision costs. The distances to some key areas are also highlighted in Table \ref{tbl:summary}. On average, households are around 13 kilometers away from an asphalt road and almost 60 kilometers away from the nearest market. This makes both the purchase of improved seed and the marketing of yields subject to transportation costs and a difficult enterprise. Fertilizer costs are on average 8,350 Birr ($\thicksim$178 USD), with a large amount of the sample not using fertilizer at all. This is particularly important in the context of improved maize seed use as these technologies are intensive in external inputs, such as chemical fertilizers and irrigation, to be effective. Only 5\% of the sample irrigates their crops making households susceptible to weather shocks.

@@

We use self-reported and cropcut yields as outcomes in a Group Random Coefficients (GRC) approach to investigate the heterogeneous effects of adoption. The advantage of the GRC approach is that we do not rely on identification from a valid instrument, but on modeling the heterogeneity directly through a projection of adoption returns on their observed ``trajectory'' of adoption. This allows us to directly capture two effects: heterogeneous effects in the form of the so-called ``comparative advantage" of adoption, and, under the assumption of linearity in comparative advantage, a description of the overall economy's sorting of farmers based on their comparative advantage. The first is important because it is able to disentangle the effect of a general household's agricultural ability (which can be thought of as a fixed effect) and their adoption-specific potential. The second parameter can tell us whether the economy sort farmers in a way that is commensurate with providing those that would benefit the most from adoption actually adopting, or whether there are barriers interfering with that. See \cite{Tjernstrom_Emilia_Dalia_Ghanem_Oscar_Barriga_Cabanillas_Travis_J_Lybbert_Jeffrey_D_Michler_and_Aleksandr_Michuda2020-bc} for more information on the assumptions involved for the GRC approach.

@@

\input{results/tables/traj}

@@

@@

@@

The GRC approach uses the variation in the adoption trajectories of each observation in the ESS 1-3 panel as a source of information that allows for an unbiased estimation. ESS4 is no longer part of this panel data-set, and so in order to estimate the effect of adoption of improved seeds on yields, an alternative estimation method is needed. We propose an endogenous switching regression model which accounts for unobserved heterogeneity in adoption decisions. This method imposes restrictive assumptions regarding the distribution of the errors for the selection and output equations. At the same time, the exclusion restriction needs to be satisfied in order to reduce the bias of the estimation.

@@

Our results use two definitions of yields: measured cropcuts of dry maize and self-reported yields maize yields. Each measure of yields has its benefits and drawbacks. Cropcuts, for instance, provide a true estimate of yields for a given parcel, but at the cost of a reduced sample size. Self-reports have larger sample sizes, but may suffer from measurement error \citep{gollin2021heterogeneity}. For controls we use the years of education of the household head, the age of the household head, the sex of the household head, whether there is a title for the land being cultivated, the parcel size in hectares, as well as the hours of household and hired labor. Commensurate with the literature, we also include interactions of these controls with improved seed adoption as added controls. Interactions with the improved seed variable are meant to further account for household specific variation that adopters may have.

@@

\input{results/tables/grc}

@@

To explore this further, we run a similar regression to calculate the comparative advantage of adopting a certain number of times. This is similar to Eq. \ref{eq:GRC_Suri}, but uses the number of times that a household adopted hybrid maize (here, $NAdopt$), not its trajectory.

In this, instead of a set of trajectories, we have a set of the number of times adopted $\mathcal{NA}$. $\nu$ denotes the average yield from adoption $na$ times. We use adopting once as the base. Figure \ref{fig:n_adopt} shows the results of the regression and calculation of comparative advantage. As expected, adopting zero times leads to negative comparative advantage, but with positive returns to adoption, implying that there is value to adopting to those that have never adopted hybrid maize seed. Adopting once leads to both positive comparative advantage, and positive returns to adoption. However, adopting twice actually leads to negative comparative advantage, and less positive returns. Adopting all three times leads to positive comparative advantage, but negative returns.

\begin{figure}
    \centering
    \caption{Aggregation Results: Comparative Advantage and Returns by the Number of Times Adopted (Dry Cropcuts)}
    \label{fig:n_adopt}
    \includegraphics[scale=0.75]{results/figures/num_adoption.png}
    \vspace*{-2em}
    \begin{table}[H]
        \centering
        \begin{tabular}{p{0.75\textwidth}} 
            \begin{tablenotes}
                  \small
                  \item Note: Author's calculations using ESS. The figures shows the aggregate impact from Table \ref{tbl:unres} using the number of time a person adopted as the basis for the aggregation.
            \end{tablenotes}
        \end{tabular}
    \end{table}
\end{figure}

@@

To complement the results of the GRC model, we present the Endogenous Switching Regression results. The fourth wave of the ESS data allows us to use DNA-fingerprinting results. This is advantageous over waves 1-3, because we do not need to rely on self-reported adoption status. We first examine a self-reported definition of an improved variety to validate the results from the GRC model. As before, we consider a looser definition of improved seeds, which includes new seeds but also second-generation and recycled (SR1), and a more strict definition, which includes only new varieties (SR2). Table \ref{tab:switch1} shows the results of the Switching regression using both definitions, and using self-reported and cropcut yields. We see that regardless of the definition of improved seed, yield measurement and controls included, the results consistently show that the ATT is negative and the ATU is positive, meaning that farmers that adopted a self-reported improved seed would've been better off not adopting. Farmers that did not adopt improved seeds would have been better off adopting. This is in line with the GRC results as it seems to imply that households are not making profit-maximizing decisions; their adoption decisions seem to be leading to a reduction in welfare, whether they are adopting or not adopting.

@@

>>>>

    \clearpage\maketitle
	\thispagestyle{empty}
	\vspace*{-2em}
    \begin{center}\begin{abstract}
    \singlespacing
			\noindent  
			Despite large investments in  developing new  maize varieties, evidence shows that few farmers consistently adopt this new technology over time. We study the adoption trajectories of farmers over a three-round panel data set to explain low adoption rates with heterogeneous returns to adoption. By relying on self-reported use of improved seeds, we find a puzzling answer for why farmers do not use new seed varieties: adoption leads to seemingly small and negative returns. Heterogeneity analysis across different levels of comparative advantage show that farmers who adopt do not have a comparative advantage in adopting improved seed, nor do they exhibit high returns to adopting. To reconcile this result with the expected higher yields for improved seeds, we exploit unique information on maize DNA-fingerprinting collected over the same areas in 2018. We find that the negative returns to adoption mask heterogeneity in the actual quality of the genetic material of the seeds farmers categorized as improved varieties in the self-reported data. Positive returns to adoption are found for those farmers that use higher-purity germplasm, drought-tolerant maize, and newly released varieties. When we account for misclassification in our heterogeneity analysis, we find that the puzzle is resolved. Our findings point to the wide dispersion of older and genetically diluted varieties, for which poorer farmers may be paying a premium. The implications of our findings speak to the need for policies to better target context and geography, expand accessibility of improved seeds, and make higher yields varieties more inclusive.

@@

        {\small \noindent\emph{JEL Classification}: O12, O13, O33, Q12, C13 \\
        	\emph{Keywords}: Improved maize, DNA-fingerprinting, Group Random Coefficient, Endogenous Switching Regression, Heterogeneous returns}

@@

%%% Why it is important
Food security is a major challenge in many East African countries, and Ethiopia is no exception, having historically struggled to provide an adequate and reliable food supply \citep{Ramakrishna2002-hv, Jaleta2018-oj}. Since the drought of 1984, Ethiopia has become increasingly reliant on maize, and as of today, maize constitutes the most widespread crop in the country, with total area cultivated increasing and average yields almost doubling since 1990.\footnote{FAO statistics show the total area cultivated with Maize in 2019 is 25\% larger than the next most widespread crop (sorghum), and 27\% and 139\% larger than the next two following crops, wheat and barley, respectively.}\footnote{Improved security conditions and maize productivity contributed with an important decrease in undernourishment which fell from 47\% in 1990 to 14\% in 2018 (The World Bank, 2021). But despite these gains, food insecurity still affects more than half the population of Ethiopia. The country has also recently seen increases in food insecurity because of COVID-19, the conflict in Ethiopia’s Tigray region and the desert locust outbreak.} But despite this, progress a large productivity gap persists between the African continent and the rest of the developing world.\footnote{From 2016 to the present, cereal crop yields in Sub-Saharan Africa have consistently been around 70-80\% lower than Europe or North America \citep{worldbankyield}.} The persistence of this gap is often attributed to different adoption rates of green revolution technologies, which include improved seeds. Evidence has shown that the adoption of improved seed leads to increases in agricultural yields \citep{Carter2014-fm}, food security \citep{Shiferaw2014-op}, and poverty reduction \citep{Minten2008-tj}.\footnote{Nevertheless, critics of green revolution technologies have also claimed that these technological packages have not been as profitable for many farmers in Africa as promised. Even the most successful of these countries, Ethiopia, is not on track to meet the goals of the Alliance for a Green Revolution for Africa (AGRA). This is due to high costs, and heavy dependence on subsidies \citep{Wise20}.} But in fact, only 13\% of maize farmers in Ethiopia have adopted improved seeds consistently between 2011 and 2015. A large percentage has adopted new varieties inconsistently, and the largest share of farmers (62\%) have not used improved seeds at all during the same period.

%old version
% \footnote{FAO statistics show the total area cultivated with Maize in 2019 is 25\% larger than the next most widespread crop (sorghum), and 27\% and 139\% larger than the next two following crops, wheat and barley, respectively.} There was an upward trend in the total area cultivated with maize since 1990, with average yields almost doubling during this period.\footnote{Improved security conditions and maize productivity contributed with an important decrease in undernourishment which fell from 47\% in 1990 to 14\% in 2018 (The World Bank, 2021). But despite these positive gains in total area planted and average yields, food insecurity still affects more than half the population of Ethiopia.The country has also recently seen increases in food insecurity because of COVID-19, the conflict in Ethiopia’s Tigray region and the desert locust outbreak.} Despite progress a large productivity gap persists between the African continent and the rest of the developing world. From 2016 to the present, cereal crop yields in Sub-Saharan Africa have consistently been around 70-80\% lower than Europe or North America \citep{worldbankyield}. The persistence of this gap is often attributed to different adoption rates of green revolution technologies, which include improved seeds. Evidence has shown that the adoption of improved seed leads to increases in agricultural yields \citep{Carter2014-fm}, food security \citep{Shiferaw2014-op}, and poverty reduction \citep{Minten2008-tj}.\footnote{Nevertheless, critics of green revolution technologies have also claimed that these technological packages have not been as profitable for many farmers in Africa as promised. Even the most successful of these countries, Ethiopia, is not on track to meet the goals of the Alliance for a Green Revolution for Africa (AGRA). This is due to high costs, and heavy dependence on subsidies \citep{Wise20}.}

@@

Several answers have been proposed to explain such low adoption of improved varieties. These include lack of liquidity and other economic constraints \citep{Carter2014-fm}, growing conditions such as the type of soil, input use, and farmer characteristics \citep{Munshi2004-og}, imperfections in credit markets \citep{Croppenstedt2003-pq}, ambiguously defined property rights \citep{Place2000-el}, lack of commitment devices \citep{Duflo2009-iv}, high transportation costs that increase the cost of agricultural inputs \citep{Byerlee2013-qk}, and improved seed supply constraints \citep{Bird2020-nt}. An alternative mechanism for a lack of improvement has the misclassification of seed and seed dilution. Net returns to improved seed is often lower than expected in the long term because a large percentage of farmers use diluted improved seed varieties, or ones that have been recycled over many generations. Evidence suggests that yield improvement and disease resistance wanes over time \citep{Abate2015-rj}. Conversely, given the ubiquity of improved seed varieties over the course of decades of access in Ethiopia, some farmers may use improved seed without knowing it, leading to inferior yields than conventional seed due to a lack of the application of complementary inputs.

Misclassification of improved seed has been studied in the case of cassava in Nigeria \citep{wossen2019estimating,macours2019comment}. The authors use DNA fingerprinting data to show that there is a 22\% bias in the productivity impact of improved cassava seed. In the context of seed adoption, however, there is evidence to suggest that the productivity impacts of seed affects households heterogeneously \citep{Suri2011-oi}, and that knowing average impacts to adoption may not be as informative. The observed effect of adoption may be low because households that would have a comparative advantage in adoption of improved seed are constrained by a lack of access or infrastructure. Additionally, the non-classical measurement error introducted by misclassification can bias heterogeneity in ways that may attenuate or even reverse productivity impacts. This paper provides a way to account for misclassification of seed when additional, error-free data exists, such as DNA fingerprinting data. We estimate the comparative advantage of households directly using a Group Random Coefficients (GRC) Model and correct its heterogeneity estimates using misclassification rates from DNA fingerprinting data of the same plots. \citep{Suri2011-oi, Tjernstrom_Emilia_Dalia_Ghanem_Oscar_Barriga_Cabanillas_Travis_J_Lybbert_Jeffrey_D_Michler_and_Aleksandr_Michuda2020-bc}.

@@

@@

Our results contribute to the literature on the drivers of low adoption rates of improved seed varieties. The consensus in this literature is that the success of improved varieties depends on a range of factors, and that the widespread adoption of  improved seeds is the result of complementary conditions and inputs that allow farmers to extract the full potential of these varieties. Specifically, extension services and on-farm field trials, seed variety characteristics and rainfall have been found to be crucial to the adoption of improved maize in Tanzania \citep{Kaliba2000-jh}. In the specific case of Ethiopia, labor, fertilizer use, farmers’ experience with extension packages, rainfall suitability and prices have been shown to increase the opportunity cost of not adopting improved wheat varieties \citep{Wale2006-bv}. Market access and the accessibility of extension services are also critical in the adoption of improved chickpea varieties in Ethiopia \citep{Verkaart2019-ol}. Other important determinants of heterogeneous effects include risk aversion \citep{Holden2016-vy}, farm size \citep{Ghimire2015-bd}, and credit constraints \citep{Simtowe2008-jn,Balana2020-hx}. We are the first to consider how heterogeneous impacts to seed adoption are affected by misclassification of seed.

% This implies that poorer farmers are paying a premium for genetically diluted hybrids, which has far-reaching policy implications about the need to foster a more inclusive strategy when making the best performing seeds available to all farmers

Our approach is divided into two parts. We start by using the first three survey rounds of the Ethiopian Socioeconomic Survey (ESS), which is a panel  covering  2011, 2013, and 2015. For each round, farmers self-reported on the seed variety they used (improved or traditional). To illustrate the effects of misclassification, we show the results of the GRC model for each ``adoption trajectory'' which is a sub-group characterized by the sequence of adoption choices made by the household. We find counter-intuitive results that seemingly show that households that adopt do not have a comparative advantage in adoption, or that households that adopt more often are less advantaged than those that adopt less. Returns to adoption are lowest for those that adopt more often.

Our second set of results use the fourth round of the ESS (conducted in 2018), which collected DNA fingerprints of harvested cropcuts in selected plots and compared its genetic material with a reference library of available varieties in Ethiopia, hence, providing a unique opportunity to understand why improved seed varieties seem not to increase returns when adopted. These data provides critical information not only about the seed types but also about their purity, sources, and their year of release. We exploit this uniquely rich information to compare returns of verifiable improved seed varieties against self-reported use of new, second generation or recycled varieties. For this, we leverage different purity thresholds (or degrees of no-contamination with other varieties), year of release, types of varieties, and their sources. Since many categorizations for DNA-fingerprinting exist, we implement an Endogenous Switching Regression (ESR) approach, to understand each categories' impact on adoption.\footnote{ESR is a standard estimation strategy to study the question of adoption with cross-sectional data (\citealt{Marenya2020-kb}, \citealt{Shiferaw2014-op}), given its capacity to account for unobserved heterogeneity in adoption.} The results of the ESR show negative effects of adopting improved seeds. But the comparison of different improved seed definitions using the DNA fingerprinting shows that when the definition of improved seed is based on greater purity of the germplasm (95\% of purity or more), returns to adoption becomes positive and significant. We obtain similar results when evaluating the specific effects of adopting drought-tolerant maize varieties, CGIAR or Exotic varieties, or more recently released varieties.

Finally, to integrate our results together, we use the misclassification rates from the fourth round of the data and re-estimate the GRC results to be robust to misclassification \cite{michuda2021three}. We find that once we account for misclassification, heterogeneity in adoption becomes similar to other contexts: higher comparative advantage is associated with lower returns, pointing to significant barriers to adoption for those that would benefit the most from it.

@@

% We contribute to this literature by exploring heterogeneous comparative advantages to adoption in Ethiopia between 2011 to 2015. %, controlling for important drivers of adoption such as rainfall (found to be the main driver of heterogeneous effects for adopting agronomic packages \citealt{Marenya2020-kb}, \citealt{Katengeza2019-af}), labor and input use. 
% Our findings provide an alternative hypothesis to the puzzle of low maize adoption, namely that the wide dispersion of older generations and recycled improved seeds may be resulting in lower yields and altering farmers' input use. The null and negative effects that we see in the GRC results are actually stemming from misclassification in self-reported adoption status. This paper is the first to take seriously the bias that this can create when estimating the treatment effect of adoption, both in a homogeneous and heterogeneous setting. This also implies a methodological discovery when investigating hybrid seed adoption: the absence of a true control group to use for analysis. With more time and more genetic dilution, it becomes more likely that even those that do report hybrid use are still likely using some level of hybrid.

@@

There were significant differences between the analysis carried out with the panel survey rounds (ESS 1-3) and the most recent survey round (ESS4). Most notably, the ESS4 is no longer a panel dataset made up of the households of ESS 1-3, but it includes important unique information on the DNA fingerprinting of seed type, which can be used to calculate misclassification rates. This is important information as the first three waves of ESS only included self-reported hybrid maize status. Since our estimation strategy to calculate returns to adoption requires a balanced panel dataset of maize growers, our main analysis includes the first three waves of the ESS. We use the fourth wave to provide further light on the low levels of returns from adoption we find in the first three ESS rounds, and to explore the impact of specific types of improved seeds.

@@

Table \ref{tbl:summary} shows a set of summary statistics for the sample. The mean parcel size for a household is about 0.15 Ha, with a maximum size of less than 3.7 Ha. Harvest labor is disproportionately skewed to family participation with around 20 days of harvest labor being conducted by family members and only about 2 days for hired labor. This is indicative of either malfunctioning labor markets, income constraints or high supervision costs. The distances to some key areas are also highlighted in Table \ref{tbl:summary}. On average, households are around 13 kilometers away from an asphalt road and almost 60 kilometers away from the nearest market. This makes both the purchase of improved seed and the marketing of yields subject to transportation costs and a difficult enterprise. Fertilizer costs are on average 8,350 Birr ($\thicksim$178 USD), with a large amount of the sample not using fertilizer at all. This is particularly important in the context of improved maize seed use as these technologies are intensive in external inputs, such as chemical fertilizers and irrigation, to be effective. Only 5\% of the sample irrigates their crops making households susceptible to weather shocks.

@@

We use a Group Random Coefficients (GRC) approach to investigate the heterogeneous effects of adoption. The advantage of the GRC approach is that we do not rely on identification from a valid instrument, but on modeling the heterogeneity directly through a projection of adoption returns on their observed ``trajectory'' of adoption. This allows us to directly capture two effects: heterogeneous effects in the form of the so-called ``comparative advantage" of adoption, and, under the assumption of linearity in comparative advantage, a description of the overall economy's sorting of farmers based on their comparative advantage. The first is important because it is able to disentangle the effect of a general household's agricultural ability (which can be thought of as a fixed effect) and their adoption-specific potential. The second parameter can tell us whether the economy sorts farmers in a way that is commensurate with providing those that would benefit the most from adoption actually adopting, or whether there are barriers interfering with that. See \cite{Tjernstrom_Emilia_Dalia_Ghanem_Oscar_Barriga_Cabanillas_Travis_J_Lybbert_Jeffrey_D_Michler_and_Aleksandr_Michuda2020-bc} for more information on the assumptions involved for the GRC approach.

@@

\afterpage{{\footnotesize\tabcolsep=1pt  % hold it local
    \centering
    \input{results/tables/traj}
    \vspace*{-4em}
    \begin{table}[H]
        \centering
        \begin{tabular}{p{0.7\textwidth}} 
            \begin{tablenotes}
                  \small
                  \item Note: Author's calculations using ESS. Maize farmers only. Table shows household adoption trajectories, their frequency and their share in the sample. 1 denotes adoption of improved maize seed, 0 otherwise. The first digit denotes adopting in wave 1, the second in wave 2, and the third in wave 3. 
            \end{tablenotes}
        \end{tabular}
    \end{table}  
    \newpage
}}

@@

Our results use two definitions of yields: measured cropcuts of dry maize and self-reported maize yields. Each measure of yields has its benefits and drawbacks. Cropcuts, for instance, provide a true estimate of yields for a given parcel, but at the cost of a reduced sample size. Self-reports have larger sample sizes, but may suffer from measurement error \citep{gollin2021heterogeneity}. For controls we use the years of education of the household head, the age of the household head, the sex of the household head, whether there is a title for the land being cultivated, the parcel size in hectares, as well as the hours of household and hired labor. Commensurate with the literature, we also include interactions of these controls with improved seed adoption as added controls. Interactions with the improved seed variable are meant to further account for household specific variation that adopters may have.

@@

To compare the results when using a dichotomous adoption decision with an intensity of adoption variable, we run a similar regression to the above to calculate the comparative advantage of adopting a certain number of times. We estimate an equation similar to Eq. \ref{eq:GRC_Suri}, but use the number of times that a household adopted an improved maize variety (here, $NAdopt$), not its trajectory.

Here instead of a set of trajectories, we have a set of the number of times adopted $\mathcal{NA}$. $\nu$ denotes the average yield from adoption $na$ times. We use adopting once as the base.

@@

The GRC approach uses the variation in the adoption trajectories of each household in the ESS 1-3 panel as a source of information that allows for an unbiased estimation. However, because it only includes self-reported use of improved seed, estimates may suffer from missclassification bias. We hence use the  DNA-fingerprinting available in ESS4 to explore possible missclassification dynamics that may better inform the results from ESS 1-3. We propose an endogenous switching regression model to account for unobserved heterogeneity in adoption decisions. This method imposes restrictive assumptions regarding the distribution of the errors for the selection (into adoption) and output equations. At the same time, the exclusion restriction needs to be satisfied in order to reduce the bias of the estimation.

@@

@@

@@

% We then estimate the GRC equation replacing the binary adoption variable by the number of times a farmer adopted an improved variety. Figure \ref{fig:n_adopt} shows the results and the calculation of the comparative advantage. As expected, adopting zero times leads to negative comparative advantage, but with positive returns to adoption, implying that there is value to adopting to those that have never adopted hybrid maize seed. Adopting once leads to both positive comparative advantage, and positive returns to adoption. However, adopting twice actually leads to negative comparative advantage, and less positive returns. Adopting all three times leads to positive comparative advantage, but negative returns.

% \begin{figure}
%     \centering
%     \caption{Aggregation Results: Comparative Advantage and Returns by the Number of Times Adopted (Dry Cropcuts)}
%     \label{fig:n_adopt}
%     \includegraphics[scale=0.75]{results/figures/num_adoption.png}
%     \vspace*{-2em}
%     \begin{table}[H]
%         \centering
%         \begin{tabular}{p{0.75\textwidth}} 
%             \begin{tablenotes}
%                   \small
%                   \item Note: Author's calculations using ESS. The figures shows the aggregate impact from Table \ref{tbl:unres} using the number of time a person adopted as the basis for the aggregation.
%             \end{tablenotes}
%         \end{tabular}
%     \end{table}
% \end{figure}

@@

To complement the results of the GRC model, we present the Endogenous Switching Regression results. We first examine a self-reported definition of an improved variety to validate the results from the GRC model. As before, we consider a looser definition of improved seeds, which includes new seeds but also second-generation and recycled (SR1), and a more strict definition, which includes only new varieties (SR2). Table \ref{tab:switch1} shows the results of the Switching regression using both definitions, and using self-reported and cropcut yields. We see that regardless of the definition of improved seed, yield measurement and controls included, the results consistently show that the ATT is negative and the ATU is positive, meaning that farmers that adopted a self-reported improved seed would've been better off not adopting. Farmers that did not adopt improved seeds would have been better off adopting. This is in line with the GRC results as it seems to imply that households are not making profit-maximizing decisions; their adoption decisions seem to be leading to a reduction in welfare, whether they are adopting or not adopting.

@@

\appendix

\subsection{GRC Results Table}

\begin{landscape}
\input{results/tables/grc}
\end{landscape}
